Explicit Scale Simulation for analysis  of RNA-sequencing  with ALDEx2

### 2024-08-22 go3.Rmd
with Justin's suggestions
Send back to Justin for comment

---
title: "Normalizations are not what you think; Explicit Scale Simulation in ALDEx2
"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor
    affiliation: Department of Biochemistry, University of Western Ontario
    email: ggloor@uwo.ca
  - name: Michelle Pistner Nixon
    affiliation: College of Information Sciences and Technology, Pennsylvania	 State University
  - name: Justin Silverman
    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania	 State University
citeproc: TRUE
cite-method: citeproc
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl
pdf-engine: latex
output:
  pdf_document 
header-includes:
  - \usepackage{setspace}
package: ALDEx2
abstract: In  high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (e.g., total size, microbial load, or total gene expression) of the underlying biological systems. Typically a statistical normalization is used  to remove unwanted technical variation in the data or the parameters of the model to enable analyses that are sensitive to scale; e.g., differential abundance and differential expression analyses. Recently we showed that all normalizations make implicit assumptions about the unmeasured system scale and that errors in these assumptions can lead to dramatic increases in false positive and false negative rates. Here we describe updates to the ALDEx2 R package that mitigate these problems by directly modeling uncertainty in the unmeasured system scale through the use of a \textit{scale model}. Scale models generalize the idea of normalizations and can be thought of as explicitly modeling the error in normalization by examining a distribution over all possible  normalizations.  Beyond enhancing the robustness of HTS analyses,  the use of scale models within ALDEx2 enhances the transparency and  reproducibility of analyses by making implicit normalizing assumptions an explicit part of the model building process.

@@

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to the flow cell. Each of these steps is a 
compositional sampling step as only a fixed-size subsample of nucleic acid is 
carried over to subsequent steps. Thus, with
each sampling step the connection between the size of the sampled DNA
pool and the scale (e.g., size, microbial load, or total gene
expression) of the measured biological system is degraded or lost.
Increasingly, researchers are turning to modified experimental protocols
(e.g., DNA spike-ins or cell counting) in an attempt to recover the lost
biological variation in scale; see for example  [@Vandeputte:2017aa;@Props:2017aa]. 
However, these protocols often do not recover
meaningful biological variation but instead recover scale variation at
the intermediate step in the sample preparation protocol where the DNA spike-in was added. 
In short, the disconnect
between sample-to-sample variation in sequencing depth and biological
variation in scale remains an outstanding challenge.

The analysis of  HTS data suffers from several known problems that can be
traced, in whole or in part, to  misspecification of scale. The first issue 
is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;Nearing:2022aa;@Li:2022aa], 
  exhibited as dataset-dependent 
FDR control.  The FDR problem is connected to the double-filtering approach
that is often used, but which is known not to have appropriate FDR control 
[@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second issue is poor performance when 
analyzing asymmetric data; data that arises because of a bias in the direction of change.
This type of data frequently arises in in-vitro selection experiments (SELEX), 
transcriptome analysis, and microbiome analysis [@Wu2021]. The third issue is that these
problems become more pronounced as more samples are collected; that is,
more information results in a worsening of the accuracy of the analysis [@nixon2023scale,@Nixon2024B]. 
The final problem is that these data are relative data, i.e. compositional [@lovell:2011;@Quinn:2019aa], 
and substantial effort has been put into removing this constraint.

The first three problems were recently shown by Nixon and Silverman [@nixon2023scale] to be a result of a mismatch between the underlying size or scale of the system and the assumptions of the normalizations used for the analysis of HTS.  Biological variation in scale often represents an important unmeasured confounder in HTS analyses [@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria  have different growth rates and RNA contents under different conditions, which  affect our ability to identify truly differentially abundant genes [@Scott:2010;@Yoshikawa:2011aa;@Lin:2018aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of any of these systems includes both relative change (composition) and absolute abundance (scale).  Current methods access only the compositional information yet make implicit assumptions about the scale.

Recently, Nixon et al. [-@nixon2023scale] showed that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models. They showed that \emph{all} normalizations make some assumption about  scale. These implicit assumptions are often  difficult to interpret, and different normalizations provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa]. Intuitively, normalizations in widespread use  assume that either all samples have the same scale, e.g. proportions, rarefaction [@Hughes:2005tu], RPKM [@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample can be chosen as a reference to which the others are scaled e.g. the TMM [@Robinson:2010a], or LVHA [@Wu2021] or the additive log-ratio [@aitchison1982]; or that different sub-parts of each sample maintain a constant scale across samples e.g. the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR  and its derivatives.

The original ALDEx2 [@fernandes:2013] model unwittingly made a strict assumption about scale through the CLR normalization [@nixon2023scale]. While this assumption could be useful in many cases it could  never be exactly true, and others have shown that it is not always the best option [@Yerke:2024aa]. Nixon et al. [-@nixon2023scale] showed that better scale assumptions resulted in more reproducible data analysis including better control of both false positive and false negative results. In essence ALDEx2 has been modified to explicitly model the scale over a range of reasonable normalization parameters. Here, we briefly introduce these modifications and then show how scale uncertainty can greatly improve modeling in transcriptome and meta-transcriptome datasets to provide more robust and reproducible results.

@@

To be  concrete, we let \(\mathbf{Y}\) denote the \emph{measured} \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).  Likewise, we can denote \(\mathbf{W}\) as the \emph{true}  amount of class \(d\) in the biological system from which sample \(n\) was obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions). That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements \(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\). Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).

All  the normalizations in current use are data transformations that can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation. We use the \(\hat{{\mathbf{W}}}\) notation to indicate that the normalization is attempting to provide an estimate of the true value.  The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al. 2011).

@@

The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose toolbox for Bayesian modeling of HTS data. For brevity, we discuss ALDEx2 in its simplest form as a tool for estimating the magnitude and statistical significance of Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models. At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{\hat{W}}\) given the estimate \(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses. For more details on ALDEx2 see [@fernandes:2013;@nixon2023scale].

By default, ALDEx2 uses the CLR normalization which can be written as \(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\) where \(G_{n}\) denotes the geometric mean of the vector \((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\). The CLR normalization equates to an implicit assumption that \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)[@nixon2023scale;@Nixon2024B] and even slight errors in this assumption introduces bias into the LFC estimate that is not accounted for when estimating uncertainty (e.g., confidence intervals or credible sets). In fact, Nixon et al. [-@nixon2023scale] showed that the only way in which the ALDEx2 model, or any normalization-based model, could ever be calibrated (e.g., control Type-I Error rates) was if this assumption was exactly true.  In support of this assertion it is known that false discovery rates vary widely by analysis, dataset, and normalization method [@hawinkel2017;@Li:2022aa,Nixon2024B]. To fix this, Nixon et al. [-@nixon2023scale] showed that models should incorporate potential error in the assumptions implied by normalizations.

Nixon et al. (2023) generalized the concept of normalizations by introducing  the concept of a \textit{scale model} to account for potential error. Scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which they called a \textit{Scale Simulation Random   Variable} (SSRV). They did this by including a model for \(\mathbf{\hat{W}}^{\perp}_{n}\). Since the CLR normalization makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be generalized by considering probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For example, the following scale model generalizes the CLR:

\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]

where \(\gamma\) is a tunable parameter drawn from a Gaussian distribution that controls the degree of uncertainty in the CLR assumption and \(x_{n}\) denotes a binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). We have made those modifications a permanent fixture of ALDEx2 which now represents the first software package designed for SSRV-based inference.

@@

It is standard practice in many fields of HTS, but especially transcriptomics, to use the dual cutoff approach graphically exemplified by volcano plots [@Cui:2003aa;@Schurch:2016aa] because not all statistically significant differences are biologically relevant. Paraphrasing this we can say that in in some types of datasets more samples leads to a majority of features being statistically significant. A detailed reasoning for this using modelled data is explained in two reports by Nixon et al. [-@nixon2023scale;-@Nixon2024B], which shows that scaled models completely address this analytic issue.

We use the data of Gierliński et al. [-@Gierlinski:2015aa] who conducted a highly replicated yeast transcriptome experiment comparing a wild-type strain with a snf2 gene knockout, \(\Delta\)snf2. This dataset has been used to argue that a dual cutoff approach is appropriate to limit the number of significant parts and for purposes of reproducibility [@Schurch:2016aa]. However, in this benchmarking study,  the number of significantly different transcripts varied between 65% to >80% of all transcripts depending on the tool; in essence the desired behavior was that almost all transcripts were significantly different. The guidance on reproducibility runs counter to standard statistical practice where power is intrinsically linked to sample size and very large sample sizes are indeed desired [@Halsey:2015aa]. Furthermore, the dual-cutoff approach is  known not to provide appropriate FDR control [@Zhang:2009aa; @Ebrahimpoor:2021aa].  Schurch et al. -@Schurch:2016aa further suggested that different tools might be better in some conditions or datasets than others because each tool has different intrinsic statistical power and Type 1 and Type 2 errors. Through the lens of scale uncertainty, the behaviour of the tools in this study show unacknowledged bias; false confidence in the precision of the estimate as sample size increases because o mis-specification  in the scale of the data. This certainty is driven by the assumptions of the tool not the actual experiment being investigated [@Nixon2024B].

@@

Using either DESeq2 or ALDEx2, a majority of transcripts are statistically significantly different between groups  with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.05; i.e. `r length(sig.des)` (`r round(length(sig.des)/nrow(yst.all), 2)*100`%, DESeq2) or `r length(sig.ald)` (`r round(length(sig.ald)/nrow(yst.all), 2)*100`%, ALDEx2) of the `r nrow(yst.all)` transcripts. Such  large numbers of significant transcripts seems biologically unrealistic  and furthermore  breaks the necessary assumption made by the normalization methods that the majority of the features must be invariant. That `r length(setdiff(sig.ald, sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2 identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does not, suggests that the choice of normalization plays a role in which results are returned as significant and that some if not the majority are driven by technical differences in the analysis [@Soneson:2013;@maza2013;@Dillies:2013;@Weiss:2017aa; @Schurch:2016aa].

```{r plot1, echo=F, fig.dim=c(5,6), fig.cap="Effect and volcano plots for unscaled and scaled transcriptome analysis. ALDEx2 was used to conduct a differential abundance (DA) analysis on the yeast transcriptome dataset. The results were plotted to show the relationship between difference and dispersion using effect plots ) or difference and the Benjamini-Hochberg corrected p-values (volcano plot). Panels A,C are for the unscaled analysis, and Panels B,D are for the scaled analysis. Each point represents the values for one transcript, with the color indicating if that transcript was significant in the scaled analysis and unscaled analysis (red) or in the unscaled analysis only (orange). Points in grey are not statistically signficantly different with any analysis. The horizontal dashed lines represent a log2(difference) of \\(\\pm 1.4\\)."}

par(mfrow=c(2,2))

@@

plot(yst.all$diff.win, yst.all$diff.btw, col=rgb(0,0,0,0.1), 
   xlab='Dispersion', ylab='log2 Difference')
title(expression(paste("A: Effect: ", gamma, " = 0")), adj=0, line= 0.8)
points(yst.all$diff.win[sig.ald], yst.all$diff.btw[sig.ald], 
  col=rgb(1,.66,0,0.3), cex=0.5, pch=19)
	abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
points(yst.all$diff.win[sig.s.ald], yst.all$diff.btw[sig.s.ald], 
  col=rgb(1,0,0,0.6), cex=0.5, pch=19)
	abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')

>>>>

---
title: "Explicit Scale Simulation for analysis  of RNA-sequencing  with ALDEx2"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor
    affiliation: Department of Biochemistry, University of Western Ontario
    email: ggloor@uwo.ca
  - name: Michelle Pistner Nixon
    affiliation: College of Information Sciences and Technology, Pennsylvania	 State University
  - name: Justin Silverman
    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania	 State University
citeproc: TRUE
cite-method: citeproc
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl
pdf-engine: latex
output:
  pdf_document 
header-includes:
  - \usepackage{setspace}
package: ALDEx2
abstract: In  high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (e.g., total size, microbial load, or total mRNA expression) of the underlying biological systems. Typically a statistical normalization is used  to remove unwanted technical variation in the data or the parameters of the model to enable analyses that are sensitive to scale; e.g., differential abundance and differential expression analyses. We recently showed that all normalizations make implicit assumptions about the unmeasured system scale and that errors in these assumptions can dramatically increase false positive and false negative rates. We demonstrated that these errors can be mitigated by accounting for uncertainty about scale using a *scale model*, which we integrated into the ALDEx2 R package. This article provides new insights into those methods, focusing on the application to transcriptomic analysis. Here we provide transcriptomic case studies demonstrating how scale models, rather than traditional normalizations, can reduce false positive and false negative rates in practice while enhancing the transparency and reproducibility of analyses. We show that these scale models replace the need for dual cutoff approaches often used to address the disconnect between practical and statistical significance. We demonstrate the utility of that scale models   built based on known housekeeping genes in complex metatranscriptomic datasets. Thus this work provides example and practical guidance on how to incorporate scale into transcriptomic analysis.

@@

\doublespacing

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to a flow cell. Each of these steps is a 
compositional sampling step as only a fixed-size subsample of nucleic acid is 
carried over to subsequent steps. 
Thus, with
each sampling step the connection between the size of the sampled DNA
pool and the scale (e.g., size, microbial load, or total gene
expression) of the measured biological system is degraded or lost. In the end, the information contained in the data relates only to relative abundances and  has an arbitrary scale imposed by the sequencing process[@lovell:2011;@Quinn:2019aa;@nixon2023scale]. 
Increasingly, researchers are turning to modified experimental protocols
 in an attempt to recover the lost
biological variation in scale; see for example  [@Vandeputte:2017aa;@Props:2017aa]. 
However, these protocols only recover scale variation at
the intermediate step in the sample preparation protocol where the intervention was made. 
In short, the disconnect
between sample-to-sample variation in sequencing depth and biological
variation in scale remains an open challenge.

The analysis of  HTS data suffers from several known problems that can be
traced, in whole or in part, to  misspecification of scale. The first issue 
is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;@Nearing:2022aa;@Li:2022aa], 
  exhibited as dataset-dependent 
FDR control and by the disconnect between statistical and biological significance[@Schurch:2016aa]. In current practice,  these issues are addressed  by a dual-filtering method, whereby both a low p-value (or equivalently a low q-value following FDR correction [@storey2003positive]) and  a large difference between groups is used to identify interesting transcripts or genes for follow-up analysis[@Schurch:2016aa]. 
This double-filtering approach is graphically exemplified by the volcano plot [@Cui:2003aa], but
 is known to not   appropriately control the  FDR  
[@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second issue is poor performance when 
analyzing data where the mean change between groups is non-zero. Such asymmetric data can arise when a gene set is expressed in one group but not the other, or when one group contains different gene content from the other. 
This type of data frequently arises in in-vitro selection experiments (SELEX), 
transcriptome analysis, and microbiome analysis [@Wu2021]. 
The third issue is that these
problems become more pronounced as more samples are collected; that is,
more information results in a worsening of the accuracy of the analysis [@Li:2022aa;@nixon2023scale,@Nixon2024B].

The  three problems were recently shown by Nixon and Silverman [@nixon2023scale] to be a result of a mismatch between the underlying size or scale of the system and the assumptions of the normalizations used for the analysis of HTS.  Biological variation in scale often represents an important unmeasured confounder in HTS analyses [@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria  have different growth rates and RNA contents under different conditions, which  affect our ability to identify truly differentially abundant genes [@Scott:2010;@Yoshikawa:2011aa;@Lin:2018aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of any of these systems includes both relative change (composition) and absolute abundance (scale).  Current methods access only the compositional information yet make implicit assumptions about the scale [@nixon2023scale,@Nixon2024B].

Recently, Nixon et al. [-@nixon2023scale] showed that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models. They showed that \emph{all} normalizations make some assumption about  scale but these implicit assumptions are often inappropriate and difficult to interpret. Because of this different normalizations provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa]. Intuitively, normalizations in widespread use  assume that either all samples have the same scale, e.g. proportions, rarefaction [@Hughes:2005tu], RPKM [@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample can be chosen as a reference to which the others are scaled e.g. the TMM [@Robinson:2010a], or LVHA [@Wu2021] or the additive log-ratio [@aitchison1982]; or that different sub-parts of each sample maintain a constant scale across samples e.g. the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR [@Aitchison:1986] and its derivatives.

The original naive ALDEx2 [@fernandes:2013] model unwittingly made a strict assumption about scale through the CLR normalization [@nixon2023scale]. While this assumption could be useful in many cases it could  never be exactly true, and others have shown that it is not always the best option [@Yerke:2024aa]. Nixon et al. [@nixon2023scale] showed that better scale assumptions resulted in more reproducible data analysis including better control of both false positive and false negative results. We recently modified ALDEx2  to explicitly model the scale over a range of reasonable normalization parameters, and showed significant improvements in performance in microbiome and in-vitro selection experiments [@Nixon2024B]. Here, we briefly introduce review modifications and then show how scale uncertainty can greatly improve modeling in transcriptome and meta-transcriptome datasets to provide more robust and reproducible results.

@@

Formal and expanded descriptions of the concepts that follow are given in [@nixon2023scale;@Nixon2024B].   To be  concrete, we let \(\mathbf{Y}\) denote the \emph{measured} \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample \(n\).  Likewise, we can denote \(\mathbf{W}\) as the \emph{true}  amount of class \(d\) in the biological system from which sample \(n\) was obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions). That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements \(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\). Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\), or as the logarithm \(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log \mathbf{W}^{\perp}_{n}\).

Many of  the normalizations in widespread use in tools such as DESeq2 [@Love:2014aa], edgeR [@Robinson:2010a], metagenomeSeq [@Paulson:2013aa] ALDEx2 [@fernandes:2014]  can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn} \approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation. We use the \(\hat{{\mathbf{W}}}\) (\(\ \hat{}\ \)) notation to indicate that the output is an estimate of the true value.  The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al. 2011).

@@

The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose toolbox to  model the uncertainty of HTS data and to use that model to estimate LFC significance.  At a high-level, ALDEx2  has three connected components to estimate the uncertainty inherent in HTS datasets. First, the tool accounts for the uncertainty of the sequencing counts using Dirichlet multinomial sampling to build a probabilistic model of the data; i.e., \(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).  Secondly, ALDEx2 uses the centred log-ratio transformation to scale the data [@fernandes:2013]. However, this step was modified recently to account for scale uncertainty and misspecification [@nixon2023scale,@Nixon2024B] via a scale model, explained  with more details in [@nixon2023scale;@Nixon2024B] and summarized in the next paragraph. Finally, a standard null-hypothesis test and a non-parametric estimate of mean standardized difference are used to report on the finite sample variation. These sources of uncertainty and variation are combined via reporting the expected values from a Monte-Carlo simulation framework. For simplicity, we use the term 'difference' to refer to the absolute difference between groups, and 'dispersion' to refer to the  within-condition difference or pooled variance as defined in [@fernandes:2013]. These are calculated on a \(\log_2\) scale. For more details on ALDEx2 see [@fernandes:2013;@fernandes:2014;@nixon2023scale;@Nixon2024B].

Scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which Nixon et al. [@nixon2023scale] term  a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon et al. [@nixon2023scale] generalized the concept of normalizations by introducing  the concept of a \textit{scale model} to account for potential error in the centred log-ratio normalization step.  They did this by including a model for \(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR method used by ALDEx2 makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where \(G_n\) is the geometric mean of sample n, which while being a random variable, is essentially constant across each Monte-Carlo replicate. With this modificaiton, ALDEx2 can be generalized by considering probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For example, the following scale model generalizes the CLR:

\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(\mu, \gamma^{2})\].

This formulation is quite flexible [@nixon2023scale;@Nixon2024B]. In the simple or 'default' configuration,   \(\mu = 0\) and  \(\gamma\) is a tunable parameter drawn from a log-Normal distribution that adds scale uncertainty.  The default approach controls only the degree of uncertainty in the CLR assumption for the \(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). In the advanced or 'informed' configuration, \(\mu\) takes different values for each group, and this controls the location of the LFC assumption. An example of both the default and informed approaches is given for a microbiome dataset in @Nixon2024B, and here we show that these approaches also work well in transcriptome and metatranscriptome datasets. These modifications are  instantiated in ALDEx2 which is  the first software package designed for SSRV-based inference.

@@

Gierliński et al. [@Gierlinski:2015aa]  generated a  highly replicated yeast transcriptome dataset to compare gene expression between a wild-type strain and a snf2 gene knockout, \(\Delta\)snf2. This dataset was  used to test several RNA-seq tools for their power  to detect the set of differentially abundant transcripts identified in the full dataset when the data was subset [@Schurch:2016aa]. In this  study each tool had its own 'gold standard' set of transcripts with different tools identifying between between 65% to >80% of all transcripts as being significantly different. Since the majority of transcripts were significantly different, the authors suggested that it was more appropriate to apply a  dual cutoff composed of both a Benjamini-Hochberg [@benjamini:1995] corrected p-value (q-value) plus a difference cutoff  to limit the number of identified transcripts to a much smaller fraction of the total.  Nixon et al, @Nixon2024B showed that adding even a small amount of  scale uncertainty with ALDEx2  dramatically reduced the number of significant transcripts identified, removing the need for the dual cutoff approach in this dataset and others. Below we include an intuitive explanation of how incorporating scale uncertainty achieved this outcome using a setting of \(\gamma=0.5\), which is in line with the recommendations of [@Nixon2024B] and that gives results comparable to those proposed in [@Schurch:2016aa].

@@

We start with the assumption that not all statistically significant differences are biologically relevant [@efron2008FDR], and that such a large number of significantly different transcripts breaks the  necessary assumption for DA/DE expression that most parts be invariant [@Dillies:2013]. As noted,  transcriptomics commonly uses a dual cutoff approach that is graphically exemplified by volcano plots [@Cui:2003aa;@Schurch:2016aa]. 
Using either DESeq2 or ALDEx2, a majority of transcripts are statistically significantly different between groups  with a  q-value cutoff of \( \le 0.05\); i.e. `r length(sig.des)` (`r round(length(sig.des)/nrow(yst.all), 2)*100`%, DESeq2) or `r length(sig.ald)` (`r round(length(sig.ald)/nrow(yst.all), 2)*100`%, ALDEx2) of the `r nrow(yst.all)` transcripts. These values are in line with those observed by [@Schurch:2016aa]. Such  large numbers of statistically significant transcripts seems biologically unrealistic. That `r length(setdiff(sig.ald, sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2 identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does not, suggests that the choice of normalization plays a role in which results are returned as significant and that some, if not the majority, are driven by technical differences in the analysis [@Dillies:2013;@Li:2022aa].

```{r plot1, echo=F, fig.dim=c(5,6), fig.cap="Volcano and effect plots for unscaled and scaled transcriptome analysis. ALDEx2 was used to conduct a differential expression (DE) analysis on the yeast transcriptome dataset. The results were plotted to show the relationship between difference and dispersion using effect plots or difference and the q-values using volcano plots. Panels A,C are for the naive analyses, and Panels B,D are for the  default analyses that include scale uncertainty. Each point represents the values for one transcript, with the color indicating if that transcript was significant in the both analyses (red) or in the naive analysis only (orange). Points in grey are not statistically signficantly different under any condition. The horizontal dashed lines represent a \\(\\log_2\\)difference of \\(\\pm 1.4\\)."}

par(mfrow=c(2,2))
# volcano
#plot(res@listData$log2FoldChange,-1*log10(res@listData$padj + 1e-300), 
#  col=rgb(0,0,0,0.1), xlab='log2 Difference', ylab='-1 log10(p.adjust)')
#title('D: DESeq2 volcano', adj=0, line= 0.8)
#points(res@listData$log2FoldChange[sig.des],-1*log10(res@listData$padj[sig.des] + 1e-300), col=rgb(1,.66,0,0.3), 
#  pch=19, cex=0.5)
#points(res@listData$log2FoldChange[sig.s.ald],-1*log10(res@listData$padj[sig.s.ald] + 1e-300), col=rgb(1,0,0,1), 
#  pch=19, cex=0.5)
#  abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')

l10q <- expression(-log[10](q))
l2d <- expression(log[2]~ Difference)

plot(yst.all$diff.btw, -1*log10(yst.all$we.eBH +1e-70), col=rgb(0,0,0,0.1), 
   xlab=l2d, ylab=l10q)
title(expression(paste("A: Volcano: ", gamma, " = 0")), adj=0, line= 0.8)
points(yst.all$diff.btw[sig.ald], -1*log10(yst.all$we.eBH[sig.ald]	+1e-70), 
  col=rgb(1,.66,0,.3), cex=0.5, pch=19)
points(yst.all$diff.btw[sig.s.ald], -1*log10(yst.all$we.eBH[sig.s.ald]	+1e-70), 
  col=rgb(1,0,0,1), cex=0.5, pch=19)
abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')

plot(yst.s.all$diff.btw, -1*log10(yst.s.all$we.eBH +1e-15), col=rgb(0,0,0,0.1),
   xlab=l2d, ylab=l10q)
title(expression(paste("B: Volcano: ", gamma, " = 0.5")), adj=0, line= 0.8)
points(yst.s.all$diff.btw[sig.s.ald], -1*log10(yst.s.all$we.eBH[sig.s.ald]+1e-15),
  col=rgb(1,0,0,1), cex=0.5, pch=19)
	abline(v=c(-1.4,1.4), lty=2,lwd=2, col='grey')

@@

plot(yst.all$diff.win, yst.all$diff.btw, col=rgb(0,0,0,0.1), 
   xlab='Dispersion', ylab=l2d)
title(expression(paste("C: Effect: ", gamma, " = 0")), adj=0, line= 0.8)
points(yst.all$diff.win[sig.ald], yst.all$diff.btw[sig.ald], 
  col=rgb(1,.66,0,0.3), cex=0.5, pch=19)
	abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
points(yst.all$diff.win[sig.s.ald], yst.all$diff.btw[sig.s.ald], 
  col=rgb(1,0,0,0.6), cex=0.5, pch=19)
	abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
